---
title: "figures"
author: "Jarning"
date: "2021/4/30"
output: html_document
---

```{r}
library(tidyverse)
library(latex2exp)
options(stringsAsFactors = FALSE)
tidyverse_conflicts()
```

```{r}
theme_set(
  theme_bw(base_size = 15) + 
    theme(axis.line = element_line(color="black"),
          axis.text = element_text(color = "black"),
          plot.title = element_text(hjust = .5, face = "bold"),
          panel.grid = element_blank(),
          legend.background = element_rect(fill=alpha('white', 0))
    )
)
```

## Figure 3

> ZFP541 negatively regulates the pre-pachytene transcriptional programs in pachytene spermatocytes

### PCA plot (3A)
```{r, fig.width=10, fig.height=4}
data.use <- read.table("~/data/data3/jarning/seq-data/z541.ss2/z541_ss2.metadata.tsv", sep = "\t")

data.use <- data.use %>% mutate(
  group = factor(group, levels = c("WT,eP","Zfp541(-/-),eP","WT,eD","Zfp541(-/-),eD")),
  group2 = factor(paste(genotype, time.point, sep = ","), levels = c("WT,260h","Zfp541(-/-),260h","WT,360h","Zfp541(-/-),360h")),
  cluster.annotation = factor(cluster.annotation, levels = c("pre-meiotic", "early pachytene", "late pachytene"))
) %>% 
  filter(cluster.annotation != "pre-meiotic")

p <- ggplot(data.use, aes(PC_1, PC_2, color = cluster.annotation)) + 
  geom_point(inherit.aes = F, data = data.use %>% dplyr::select(PC_1, PC_2), aes(PC_1, PC_2), color="grey", size = .2) + 
  geom_point(shape=16, size = 1) + 
  scale_color_manual(values = c("#A05F23","#13816E")) + 
  guides(color = guide_legend(title = "Stage", override.aes = list(size = 3))) + 
  facet_wrap(~group2, ncol = 4) + 
  theme(legend.position = "top", 
        strip.text = element_blank(),
        strip.background = element_blank())

png("../figures/Figure3A.png", width = 10, height = 4, units = "in", res = 600)
p
dev.off()
```

### MA plot (3B)

```{r}
MAPlot <- function(data.use, genes.show, title) {
  data.use <- data.use %>% 
    filter(!is.na(change)) %>% 
    mutate(my.color = ifelse(change!="NS", change, "unchange"),
           my.size = ifelse(my.color == "unchange", .2, 1),
           my.alpha = ifelse(my.color == "unchange", .2, .5),
           my.label = ifelse(gene %in% genes.show, gene, NA)) %>% 
    mutate(my.color = factor(my.color, levels = c("up in KO", "unchange", "down in KO")))
  
  data.use %>% 
    ggplot(aes(baseMean, log2FoldChange, color=my.color, fill=my.color)) + 
    geom_point(size=data.use$my.size, alpha=data.use$my.alpha) + 
    geom_point(data = . %>% filter(!is.na(my.label)), size=2, shape = 21, color="black") + 
    scale_color_manual(values = c("#F46D43","grey","#74ADD1")) + 
    scale_fill_manual(values = c("#F46D43","grey","#74ADD1")) + 
    ggrepel::geom_text_repel(data = . %>% filter(!is.na(my.label)), aes(baseMean, log2FoldChange, label=my.label), show.legend = F, inherit.aes = F, max.iter = 20000, fontface = "italic") + 
    labs(x="Mean expression level", y="log2 fold change", title=title) + 
    scale_x_log10() + 
    theme(legend.title = element_blank())
}
```


```{r, fig.width=5, fig.height=4}
data.use <- read.table("../data/WT_vs_Z5KO.lP.DEGs.ss2.tsv", sep = "\t")
data.use <- data.use %>% 
  mutate(gene = rownames(.)) %>% 
  mutate(gene = plyr::mapvalues(gene, from = c("Gm960","Ccdc36"), to = c("Top6bl", "Iho1")))

#dsb.genes <- c("Prdm9","Spo11","Gm960","Mei1","Mei4","Hormad1","Ccdc36(Iho1)","Rec114","Ankrd31")
dsb.genes <- c("Prdm9","Spo11","Top6bl","Mei1","Mei4","Hormad1","Iho1","Rec114","Ankrd31")
#meiosis.regulators <- c("Rxra","Rarg","Ncoa1","Ncoa2","Gm4969","Stra8","Meioc","Ythdc2")
p <- MAPlot(data.use, genes.show = dsb.genes, title = "")
p <- p + theme(legend.position = c(0.83,0.85))

png("../figures/Figure3B.png", width = 5, height = 4, units = "in", res = 600)
p
dev.off()
```

### 3C: DSB genes

```{r}
library(ggsignif)
library(Seurat)
```

```{r}
seu <- readRDS("~/data/data3/jarning/seq-data/z541.ss2/z541_ss2.seu_final.rds")
seu$group2 = factor(paste(seu$genotype, seu$abbr.cell_type, sep = ","), levels = c("WT,eP","Zfp541(-/-),eP","WT,lP","Zfp541(-/-),lP"))

eP.DEGs <- read.table("../data/WT_vs_Z5KO.eP.DEGs.ss2.tsv", sep = "\t")
lP.DEGs <- read.table("../data/WT_vs_Z5KO.lP.DEGs.ss2.tsv", sep = "\t")
```

```{r, fig.width=5, fig.height=4}
dsb.genes <- c("Prdm9","Gm960","Mei1","Mei4","Hormad1")

genes.use <- c(dsb.genes)
DefaultAssay(seu) <- "RNA"
data.use <- FetchData(seu, vars = c("group2", genes.use))

data.signif.1 <- data.frame(
  padj=eP.DEGs[genes.use, ]$padj, 
  gene=factor(genes.use, levels=genes.use),
  y=sapply(genes.use, function(xx) max(data.use[,xx]))
)
data.signif.1$padj <- ifelse(is.na(data.signif.1$padj), 1, data.signif.1$padj)

data.signif.2 <- data.frame(
  padj=lP.DEGs[genes.use, ]$padj, 
  gene=factor(genes.use, levels=genes.use),
  y=sapply(genes.use, function(xx) max(data.use[,xx]))
)
data.signif.2$padj <- ifelse(is.na(data.signif.2$padj), 1, data.signif.2$padj)

alias <- c("Top6bl")
names(alias) <- c("Gm960")
data.signif.1$gene <- plyr::mapvalues(x=data.signif.1$gene, from=names(alias), to = alias)
data.signif.2$gene <- plyr::mapvalues(x=data.signif.2$gene, from=names(alias), to = alias)
genes.use <- plyr::mapvalues(x=genes.use, from=names(alias), to = alias)
data.use <- data.use %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "gene", values_to = "expr") %>% 
  mutate(gene = plyr::mapvalues(x=gene, from=names(alias), to = alias)) %>% 
  mutate(gene = factor(gene, levels = genes.use))

p <- data.use %>% 
  ggplot(aes(group2, expr)) + 
  geom_violin(aes(color=group2, fill=group2), scale = "width") + geom_jitter(size = .1, alpha=.2) + 
  geom_signif(data = data.signif.1, 
              aes(y_position=y-0.3, xmin=1, xmax=2, annotations=formatC(padj, 1)), manual = T) + 
  geom_signif(data = data.signif.2, 
              aes(y_position=y-0.3, xmin=3, xmax=4, annotations=formatC(padj, 1)), manual = T) + 
  facet_wrap(~gene, scales = "free_y", nrow = 2) + 
  scale_fill_manual(values = c("#A6611A", "#DFC27D", "#018571", "#80CDC1")) + 
  scale_color_manual(values = c("#A6611A", "#DFC27D", "#018571", "#80CDC1")) + 
  guides(color=guide_legend(nrow = 2), fill=guide_legend(nrow = 2)) + 
  labs(x="", y="log(CPM/10+1)") + 
  theme(legend.position = "bottom", 
        legend.title = element_blank(),
        strip.text = element_text(face = "italic"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

pdf("../figures/Figure3C.lgd.pdf", width = 5, height = 4)
lgd <- cowplot::get_legend(p)
cowplot::plot_grid(lgd)
dev.off()

png("../figures/Figure3C.png", width = 5, height = 4, units = "in", res = 600)
p + theme(legend.position = "none")
dev.off()
```

### 3D: the meiotic program

```{r, fig.width=13, fig.height=5}
tfs <- c("Gm4969","Rxra","Rarg","Ncoa1","Ncoa2","Ythdc2")
downstream <- c("Zcwpw1", "Mcmdc2", "Rad51", "Hfm1", "Rnf212", "Msh5", "Tex15", "4930447C04Rik", "Syce3")

genes.use <- c(downstream, tfs)
DefaultAssay(seu) <- "RNA"
data.use <- FetchData(seu, vars = c("group2", genes.use))

data.signif.1 <- data.frame(
  padj=eP.DEGs[genes.use, ]$padj, 
  gene=factor(genes.use, levels=genes.use),
  y=sapply(genes.use, function(xx) max(data.use[,xx]))
)
data.signif.1$padj <- ifelse(is.na(data.signif.1$padj), 1, data.signif.1$padj)

data.signif.2 <- data.frame(
  padj=lP.DEGs[genes.use, ]$padj, 
  gene=factor(genes.use, levels=genes.use),
  y=sapply(genes.use, function(xx) max(data.use[,xx]))
)
data.signif.2$padj <- ifelse(is.na(data.signif.2$padj), 1, data.signif.2$padj)

alias <- c("Six6os1", "Meiosin")
names(alias) <- c("4930447C04Rik", "Gm4969")
data.signif.1$gene <- plyr::mapvalues(x=data.signif.1$gene, from=names(alias), to = alias)
data.signif.2$gene <- plyr::mapvalues(x=data.signif.2$gene, from=names(alias), to = alias)
genes.use <- plyr::mapvalues(x=genes.use, from=names(alias), to = alias)

p <- data.use %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "gene", values_to = "expr") %>% 
  mutate(gene = plyr::mapvalues(x=gene, from=names(alias), to = alias)) %>% 
  mutate(gene = factor(gene, levels = genes.use)) %>% 
  ggplot(aes(group2, expr)) + 
  geom_violin(aes(color=group2, fill=group2), scale = "width") + 
  geom_jitter(size = .1, alpha=.2) + 
  geom_signif(data = data.signif.1, 
              aes(y_position=y-0.2, xmin=1, xmax=2, annotations=formatC(padj, 1)), manual = T) + 
  geom_signif(data = data.signif.2, 
              aes(y_position=y-0.2, xmin=3, xmax=4, annotations=formatC(padj, 1)), manual = T) + 
  facet_wrap(~gene, scales = "free_y", nrow = 2) + 
  scale_fill_manual(values = c("#A6611A", "#DFC27D", "#018571", "#80CDC1")) + 
  scale_color_manual(values = c("#A6611A", "#DFC27D", "#018571", "#80CDC1")) + 
  labs(x="", y="log(CPM/10+1)") +
  theme(legend.position = "bottom", 
        legend.title = element_blank(),
        strip.text = element_text(face = "italic"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

png("../figures/Figure3D.png", width = 13, height = 5, units = "in", res = 600)
p + theme(legend.position = "none")
dev.off()
```


### 3E: spermatogonia genes
```{r, fig.width=4, fig.height=4}
genes.use <- c("Kit", "Uchl1", "Dnmt3b", "Chd7")

DefaultAssay(seu) <- "RNA"
data.use <- FetchData(seu, vars = c("group2", genes.use))

data.signif.1 <- data.frame(
  padj=eP.DEGs[genes.use, ]$padj, 
  gene=factor(genes.use, levels=genes.use),
  y=sapply(genes.use, function(xx) max(data.use[,xx]))
)
data.signif.1$padj <- ifelse(is.na(data.signif.1$padj), 1, data.signif.1$padj)

data.signif.2 <- data.frame(
  padj=lP.DEGs[genes.use, ]$padj, 
  gene=factor(genes.use, levels=genes.use),
  y=sapply(genes.use, function(xx) max(data.use[,xx]))
)
data.signif.2$padj <- ifelse(is.na(data.signif.2$padj), 1, data.signif.2$padj)

alias <- c("Meiosin")
names(alias) <- c("Gm4969")
data.signif.1$gene <- plyr::mapvalues(x=data.signif.1$gene, from=names(alias), to = alias)
data.signif.2$gene <- plyr::mapvalues(x=data.signif.2$gene, from=names(alias), to = alias)
genes.use <- plyr::mapvalues(x=genes.use, from=names(alias), to = alias)
data.use <- data.use %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "gene", values_to = "expr") %>% 
  mutate(gene = plyr::mapvalues(x=gene, from=names(alias), to = alias)) %>% 
  mutate(gene = factor(gene, levels = genes.use))

p <- data.use %>% 
  ggplot(aes(group2, expr)) + 
  geom_violin(aes(color=group2, fill=group2), scale = "width") + geom_jitter(size = .1, alpha=.2) + 
  geom_signif(data = data.signif.1, 
              aes(y_position=y-0.3, xmin=1, xmax=2, annotations=formatC(padj, 1)), manual = T) + 
  geom_signif(data = data.signif.2, 
              aes(y_position=y-0.3, xmin=3, xmax=4, annotations=formatC(padj, 1)), manual = T) + 
  facet_wrap(~gene, scales = "free_y", nrow = 2) + 
  scale_fill_manual(values = c("#A6611A", "#DFC27D", "#018571", "#80CDC1")) + 
  scale_color_manual(values = c("#A6611A", "#DFC27D", "#018571", "#80CDC1")) + 
  guides(color=guide_legend(nrow = 2), fill=guide_legend(nrow = 2)) + 
  labs(x="", y="log(CPM/10+1)") + 
  theme(legend.position = "bottom", 
        legend.title = element_blank(),
        strip.text = element_text(face = "italic"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

png("../figures/Figure3E.png", width = 4, height = 4, units = "in", res = 600)
p + theme(legend.position = "none")
dev.off()
```

### 3F: Signature compare (Heatmap, 10x)

```{r}
library(ComplexHeatmap)
library(circlize)
signatures <- c(
  "31P: SSC",
  "31N: Diff.Spg",
  "33N: broad Spg",
  "2P: B.Spg double peak",
  "2N: B.Spg single peak",
  "5N: (pre)Leptotene",
  "44N: Zygotene",
  "38P: MSCI",
  "27P: early General",
  "23P: Z-P",
  "13N: early Pachytene",
  "39N: late Pachytene",
  "20N: Meiotic Division"
)
```

```{r}
data.use <- read_tsv("../data/WT_vs_Z5KO.10x.gene_module.tsv")

data.2 <- data.use %>% 
  pivot_wider(id_cols = stage, names_from = "group", values_from = "t")
rn <- data.2$stage
data.2$stage <- NULL
data.2 <- as.matrix(data.2)
rownames(data.2) <- rn
data.ht <- data.2
data.ht <- data.ht[!is.na(rowSums(data.ht)), ]
colnames(data.ht) <- signatures

data.ht <- data.ht[, setdiff(colnames(data.ht), c("27P: early General","2N: B.Spg single peak","23P: Z-P"))]
data.ht <- data.ht[rownames(data.ht)[-nrow(data.ht)], ]
```

```{r, fig.width=6, fig.height=5}
col_range = c(-10,0,20)
col_fun <- colorRamp2(col_range, c("blue","white","red"))
at = seq(-10, 20, by = 5)
lgd = Legend(at = at, title = "t", border="black", legend_gp = gpar(fill = col_fun(at)))

ht <- Heatmap(data.ht, cluster_rows = F, cluster_columns = F, column_names_rot = -90,
              col = col_fun,
              name = "t",
              show_heatmap_legend = FALSE,
              cell_fun = function(j, i, x, y, width, height, fill) {
                grid.rect(x = x, y = y, width = width, height = height, gp = gpar(col = "black", fill = NA))
                grid.text(sprintf("%.1f", data.ht[i, j]), x, y, gp = gpar(fontsize = 10))
              })

pdf("../figures/Figure3F.pdf", width = 6, height = 5)
draw(ht, heatmap_legend_list = list(lgd))
dev.off()
```

## Figure 4

> ZFP541 achieves its repression role via maintaining the silenced state of pre-pachytene transcriptional programs.

```{r}
seu <- readRDS("~/data/data3/jarning/seq-data/SHIP.10X/seu.rm_doublets.addMS.rds")
cluster.annotation <- read_tsv("../../zhangxinxia_scrna/analysis_SHIPC/clusters.annotations.txt")
seu$annotation <- plyr::mapvalues(seu$seurat_clusters.reorder, cluster.annotation$cluster_index, cluster.annotation$annotation)
```

### 4A: principle curve

```{r}
p.curve <- readRDS("~/data/data3/jarning/seq-data/SHIP.10X/p.curve.rds")

get_path <- function(pc.curve, df="7"){
  df = paste0("df_", df)
  pc.curve[[df]]$s[pc.curve[[df]]$ord, ] %>% 
    as.data.frame() %>% 
    set_names("X", "Y") %>% 
    distinct(X, Y)
}

get_arrow <- function(data.path, reverse = FALSE){
  if (reverse) {
    data.arrow <- data.frame(
      X = head(data.path)[5, "X"],
      Y = head(data.path)[5, "Y"],
      Xend = head(data.path)[1, "X"],
      Yend = head(data.path)[1, "Y"]
    )
  }else{
    data.arrow <- data.frame(
      X = tail(data.path)[1, "X"],
      Y = tail(data.path)[1, "Y"],
      Xend = tail(data.path)[5, "X"],
      Yend = tail(data.path)[5, "Y"]
    )
  }
  return(data.arrow)
}
```

```{r, fig.width=5, fig.height=5}
dims = paste0("UMAP_", 1:2)
meta = c("annotation", "pc.pseudotime")
data.point <- FetchData(seu, vars = c(dims, meta))

data.path <- get_path(p.curve, df=8)
data.arrow <- get_arrow(data.path, reverse = T)
pts <- c(min(data.point$pc.pseudotime), median(data.point$pc.pseudotime), max(data.point$pc.pseudotime))

p <- ggplot() + 
  geom_point(data = data.point, aes(UMAP_1, UMAP_2, color = pc.pseudotime), size = .2) + 
  geom_path(data = data.path, aes(X,Y), size = 1) + 
  geom_segment(data = data.arrow, aes(x = X, xend = Xend, y = Y, yend = Yend), 
               arrow = arrow(length = unit(0.1, "in"), angle = 30, type = "closed"), size = 1) + 
  scale_color_continuous(breaks = pts, labels = seq(0,100,50)) + 
  guides(color = guide_colorbar(title = "Pseudotime")) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = c(0.4, 0.8))

png("../figures/Figure4A.png", width = 5, height = 5, units = "in", res = 600)
p
dev.off()
```

### 4B: Pseudotime

```{r, fig.width=6, fig.height=3}
pts <- c(min(data.point$pc.pseudotime), median(data.point$pc.pseudotime), max(data.point$pc.pseudotime))

p <- ggplot(data = data.point, aes(annotation, pc.pseudotime, fill=annotation)) + 
  geom_violin(scale = "width") + 
  ggsci::scale_fill_d3("category20") + 
  scale_y_continuous(breaks = pts, labels = seq(0,100,50)) + 
  labs(x="", y="Pseudotime") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle=45, hjust = 1))

pdf("../figures/Figure4B.pdf", width = 6, height = 3)
p  
dev.off()
```

### 4C: known markers

```{r}
seu.magic <- readRDS("../../../seq-data/SHIP.10X/seu.magic.rds")
annot.levels <- levels(seu.magic$annotation)
cluster.annotation <- read_tsv("../../zhangxinxia_scrna/analysis_SHIPC/clusters.annotations.txt")
seu.magic$annotation <- plyr::mapvalues(seu.magic$seurat_clusters.reorder, 
                                        cluster.annotation$cluster_index, cluster.annotation$annotation)
seu.magic$annotation <- factor(seu.magic$annotation, levels = annot.levels)

flattern_imputated_data <- function(data.use, bin.width = 10, min.cells = 1, value = "") {
  data.use %>%
    mutate(pseudo.bin = round(pc.pseudotime / bin.width, 0)) %>%
    group_by(pseudo.bin, genotype) %>%
    summarise(median.exp = median(get(value)),
              counts = n()) %>%
    ungroup() %>%
    mutate(pc.pseudotime = pseudo.bin * bin.width) %>%
    filter(counts >= min.cells)
}
```

```{r, fig.width=20, fig.height=4}
genes <- c("Ccnd2","Stra8","Sycp2","Piwil1")
dims = c("pc.pseudotime")
meta = c("annotation", "genotype")
data <- FetchData(seu.magic, vars = c(dims, meta, genes))

plot.list <- lapply(genes, function(xx) {
  data.use <- flattern_imputated_data(data, bin.width = 10, value = xx)
  max.val <- max(data.use$median.exp)
  pts <- c(min(data$pc.pseudotime), median(data$pc.pseudotime), max(data$pc.pseudotime))
  ggplot() +
    geom_line(data = data.use, aes(pc.pseudotime, median.exp, color=genotype, group=genotype), size = 1) +
    geom_jitter(data = data, aes(pc.pseudotime, -max.val*0.1-0.1, fill = annotation), show.legend = F,
                stroke = 0, color="white", shape=21, height = max.val*0.05, size = 1, alpha = .5) +
    ggsci::scale_fill_d3("category20") + 
    geom_vline(xintercept = -22540, linetype="dashed", color="blue") +
    scale_x_continuous(breaks = pts, labels = seq(0,100,50)) + 
    guides(color = guide_legend(title = "", override.aes = list(size = 1.5))) + 
    labs(x="Pseudotime", y="Expression level", title = xx) + 
    theme(legend.position = c(0.8, 0.95),
          plot.title = element_text(face = "italic"))
})

pdf("../figures/Figure4C.lgd.pdf", width = 5, height = 4)
lgd <- cowplot::get_legend(plot.list[[1]])
cowplot::plot_grid(lgd)
dev.off()

png("../figures/Figure4C.png", width = 20, height = 4, units = "in", res = 600)
plot.list <- lapply(plot.list, function(.x) .x + theme(legend.position = "none"))
cowplot::plot_grid(plotlist = plot.list, nrow = 1)
dev.off()
```

### 4D: preleptotene genes

```{r, fig.width=20, fig.height=4}
genes <- c("Gm4969", "Zcwpw1", "Gm960", "Mei1")
alias <- c("Meiosin", "Top6bl")
names(alias) <- c("Gm4969", "Gm960")

dims = c("pc.pseudotime")
meta = c("annotation", "genotype")
data <- FetchData(seu.magic, vars = c(dims, meta, genes))

plot.list <- lapply(genes, function(xx) {
  data.use <- flattern_imputated_data(data, bin.width = 10, value = xx)
  max.val <- max(data.use$median.exp)
  pts <- c(min(data$pc.pseudotime), median(data$pc.pseudotime), max(data$pc.pseudotime))
  ggplot() +
    geom_line(data = data.use, aes(pc.pseudotime, median.exp, color=genotype, group=genotype), size = 1) +
    geom_jitter(data = data, aes(pc.pseudotime, -max.val*0.1-0.1, fill = annotation), show.legend = F,
                stroke = 0, color="white", shape=21, height = max.val*0.05, size = 1, alpha = .5) +
    ggsci::scale_fill_d3("category20") + 
    geom_vline(xintercept = -22540, linetype="dashed", color="blue") +
    scale_x_continuous(breaks = pts, labels = seq(0,100,50)) + 
    guides(color = guide_legend(title = "", override.aes = list(size = 1.5))) + 
    labs(x="Pseudotime", y="Expression level", title = ifelse(xx %in% names(alias), alias[xx], xx)) + 
    theme(legend.position = c(0.8, 0.95),
          plot.title = element_text(face = "italic"))
})

png("../figures/Figure4D.png", width = 20, height = 4, units = "in", res = 600)
plot.list <- lapply(plot.list, function(.x) .x + theme(legend.position = "none"))
cowplot::plot_grid(plotlist = plot.list, nrow = 1)
dev.off()
```


### 4E: spermatogonia genes
```{r, fig.width=20, fig.height=4}
genes <- c("Kit", "Uchl1", "Dnmt3b", "Chd7")

dims = c("pc.pseudotime")
meta = c("annotation", "genotype")
data <- FetchData(seu.magic, vars = c(dims, meta, genes))

plot.list <- lapply(genes, function(xx) {
  data.use <- flattern_imputated_data(data, bin.width = 10, value = xx)
  max.val <- max(data.use$median.exp)
  pts <- c(min(data$pc.pseudotime), median(data$pc.pseudotime), max(data$pc.pseudotime))
  ggplot() +
    geom_line(data = data.use, aes(pc.pseudotime, median.exp, color=genotype, group=genotype), size = 1) +
    geom_jitter(data = data, aes(pc.pseudotime, -max.val*0.1-0.1, fill = annotation), show.legend = F,
                stroke = 0, color="white", shape=21, height = max.val*0.05, size = 1, alpha = .5) +
    ggsci::scale_fill_d3("category20") + 
    geom_vline(xintercept = -22540, linetype="dashed", color="blue") +
    scale_x_continuous(breaks = pts, labels = seq(0,100,50)) + 
    guides(color = guide_legend(title = "", override.aes = list(size = 1.5))) + 
    labs(x="Pseudotime", y="Expression level", title = xx) + 
    theme(legend.position = c(0.8, 0.95),
          plot.title = element_text(face = "italic"))
})

png("../figures/Figure4E.png", width = 20, height = 4, units = "in", res = 600)
plot.list <- lapply(plot.list, function(.x) .x + theme(legend.position = "none"))
cowplot::plot_grid(plotlist = plot.list, nrow = 1)
dev.off()
```

## Figure 5

> ZFP541 is a transcriptional repressor binding to the promoters of meiotic prophase genes and their upstream meiosis initiators.

```{r}
library(ChIPseeker)
library(tidyverse)
library(org.Mm.eg.db)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(RColorBrewer)
library(eulerr)
options(stringsAsFactors = F)

txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

tidyverse_conflicts()
```

### 5A: ZFP541 peaks annotation

```{r}
peakannot <- function(peak, summit) {
  s1 <- readPeakFile(peak)
  s2 <- read_tsv(summit, col_names = F)
  s2 <- s2 %>% set_names(c("chr","start","end","name","log10Q")) %>% dplyr::filter(log10Q >= 2)
  s1 <- s1 %>% subset(name %in% s2$name)
  s1.anno <- annotatePeak(s1, TxDb=txdb, tssRegion=c(-3000, 3000), addFlankGeneInfo=TRUE, annoDb="org.Mm.eg.db")
  return(s1.anno)
}
```

```{r}
z5.1 <- peakannot(peak = "../data/fig.3a.peaks/Zfp541_rep1_peaks.xls", 
                  summit = "../data/fig.3a.peaks/Zfp541_rep1_summits.rmBL.bed")

z5.2 <- peakannot(peak = "../data/fig.3a.peaks/Zfp541_rep2_peaks.xls", 
                  summit = "../data/fig.3a.peaks/Zfp541_rep2_summits.rmBL.bed")

z5.ko.1 <- peakannot(peak = "../data/fig.3a.peaks/Zfp541-KO_rep1_peaks.xls", 
                     summit = "../data/fig.3a.peaks/Zfp541-KO_rep1_summits.rmBL.bed")

modify.annoStat <- function(aa, sample){
  annoStat = aa@annoStat
  annoStat$sample = sample
  annoStat$peakNum = aa@peakNum
  return(annoStat)
}

data.use <- bind_rows(modify.annoStat(z5.1, "Zfp541 rep1"), 
                      modify.annoStat(z5.2, "Zfp541 rep2"), 
                      modify.annoStat(z5.ko.1, "Zfp541-KO"))

data.use <- data.use %>% mutate(peaks = Frequency * peakNum / 100)

saveRDS(data.use, "../data/peaks.stat.rds")
write_tsv(z5.1@anno %>% as.data.frame(), "../data/fig.3a.peaks/Zfp541_rep1_peaks.annot.tsv")
write_tsv(z5.2@anno %>% as.data.frame(), "../data/fig.3a.peaks/Zfp541_rep2_peaks.annot.tsv")
write_tsv(z5.ko.1@anno %>% as.data.frame(), "../data/fig.3a.peaks/Zfp541-KO_peaks.annot.tsv")
```

```{r, fig.width=5, fig.height=4}
data.use <- readRDS("../data/fig.3a.peaks/peaks.stat.rds")

data.use <- data.use %>% filter(sample %in% c("Zfp541-KO rep1","Zfp541 rep1","Zfp541 rep2"))

data.label <- data.use %>% 
  filter(Feature=="Promoter (<=1kb)", sample %in% paste0("Zfp541 rep",1:3))

data.label2 <- data.use %>% 
  filter(Feature=="Promoter (<=1kb)")

p <- data.use %>% 
  ggplot(aes(sample, peaks, fill=Feature)) + 
  geom_col() + 
  geom_text(inherit.aes = F, data = data.label, 
            aes(sample, peaks-2000, label=paste0(sprintf("%0.1f", Frequency), "%"))) + 
  geom_text(inherit.aes = F, data = data.label2,
            aes(sample, peakNum+500, label=peakNum)) + 
  ggsci::scale_fill_d3("category20") +
  xlab("") + ggtitle("Annotation of peaks") + 
  theme_classic(base_size = 15) + 
  theme(axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = .5))

pdf("../figures/Figure5A.pdf", width = 5, height = 4)
p
dev.off()
```

### 5B: motif

```{r}
library(seqLogo)
```

```{r}
plotMotif <- function(motif.file, pdf.file){
  m <- read.table(motif.file, skip = 1)
  m <- t(m)
  m <- m/rep(colSums(m), each=4)
  motif<-makePWM(m)
  pdf(pdf.file, width = 5, height = 2)
  seqLogo(motif, xaxis=F, yaxis=F, ic.scale = F)
  dev.off()  
}

plotMotif("../data/fig.3b.motifs/rep1.motif", "../data/fig.3b.motifs/rep1.motif.pdf")
plotMotif("../data/fig.3b.motifs/rep2.motif", "../data/fig.3b.motifs/rep2.motif.pdf")
```


### 5C: ZFP541 binding genes

```{r}
get.symbol <- . %>% as_tibble() %>% 
  dplyr::filter(annotation=="Promoter (<=1kb)") %>% 
  distinct(SYMBOL) %>% 
  dplyr::filter(!is.na(SYMBOL))

z5.genes.list <- list(
  wt.rep1 = get.symbol(z5.1)$SYMBOL,
  wt.rep2 = get.symbol(z5.2)$SYMBOL,
  ko.rep1 = get.symbol(z5.ko.1)$SYMBOL
)

z5.genes.list.filter <- list(
  rep1 = setdiff(z5.genes.list$wt.rep1, z5.genes.list$ko.rep1),
  rep2 = setdiff(z5.genes.list$wt.rep2, z5.genes.list$ko.rep1)
)
```

```{r, fig.width=4, fig.height=4}
blue1 <- brewer.pal(6, "Blues")[1]
blue2 <- brewer.pal(6, "Blues")[5]

pdf("../figures/Figure3C.pdf", width = 4, height = 4)
plot(euler(z5.genes.list.filter), quantities = T, fills = c(rep(blue1,2), rep(blue2,1)), labels = T, main="ZFP541-bound genes")
dev.off()
```

```{r}
z5.genes <- UpSetR::fromList(z5.genes.list.filter)
rownames(z5.genes) <- unique(unlist(z5.genes.list.filter))
z5.genes$observed <- rowSums(z5.genes)

write.table(z5.genes, "../data/ZFP541_binding_genes.tsv", sep = "\t")
```

### 5D: ZFP541 regulated genes 
```{r}
z5.genes <- read.table("../data/ZFP541_binding_genes.tsv", sep = "\t")

# DEGs
data <- read.table("../data/WT_vs_Z5KO.lP.DEGs.ss2.tsv", sep = "\t")
upreg.genes <- subset(data, change == "up in KO") %>% rownames()
downreg.genes <- subset(data, change == "down in KO") %>% rownames()
all.genes <- rownames(data)

z5.genes$up_in_KO <- rownames(z5.genes) %in% upreg.genes %>% as.integer()
z5.genes$down_in_KO <- rownames(z5.genes) %in% downreg.genes %>% as.integer()

genes.list <- list(
  "Zfp541-bound genes" = subset(z5.genes, observed >= 2) %>% rownames(),
  "Up in KO" = upreg.genes,
  "Down in KO" = downreg.genes
)

write.table(z5.genes, "../data/ZFP541_regulated_genes.tsv", sep = "\t")
```

```{r, fig.width=4, fig.height=4}
pdf("../figures/Figure3D.pdf", width = 4, height = 4)
plot(euler(genes.list), quantities = T, fills=c("white", "#F8B4BF", "#B0E1F7"))
dev.off()
```

```{r}
ftest <- function(gene.set.1, gene.set.2, background, alternative = "greater") {
  a <- length(intersect(gene.set.1, gene.set.2))
  b <- length(setdiff(gene.set.1, gene.set.2))
  c <- length(setdiff(gene.set.2, gene.set.1))
  d <- length(setdiff(setdiff(background, gene.set.1), gene.set.2))
  fisher.test(matrix(c(a,b,c,d), ncol = 2), alternative = alternative)
}

ftest(upreg.genes, genes.list$`Zfp541-bound genes`, all.genes, "greater")
ftest(downreg.genes, genes.list$`Zfp541-bound genes`, all.genes, "greater")
```

### 5E: BETA

BETA: predict activate/repress function of Zfp541. See [reference](http://cistrome.org/BETA/tutorial.html)
Inputs:
1. Zfp541 binding sites: Zfp541_peaks.bed3
2. RNA-seq: (BSF format)

```{r}
## for BSF
data <- read.table("../data/WT_vs_Z5KO.lP.DEGs.ss2.tsv", sep = "\t")
write.table(data[c("log2FoldChange", "padj")], "../data/WT_vs_Z5KO.lP.DEGs.ss2.bsf", sep = "\t", col.names = F, quote = F)
```

```{bash}
BETA basic -p ../data/Zfp541_peaks.3c.bed \
-e ../data/WT_vs_Z5KO.lP.DEGs.ss2.bsf -k BSF -g mm10 --df 0.01 -n Zfp541_eD \
--gname2 \
--bl --bf ~/data/data1/software/BETA_1.0.7/BETA/references/mm10_CTCF_bound.bed \
-o ../figures/Zfp541_lP.basic
```


## Figure S8

> Single-cell RNA-seq via Smart-seq2 on early and late pachytene spermatocytes from WT and Zfp541-/- testis.

### S8A: pachytene markers
```{r}
seu <- readRDS("~/data/data3/jarning/seq-data/z541.ss2/z541_ss2.seu_processed.rds")
```

```{r, fig.width=6, fig.height=10}
Idents(seu) <- factor(seu$cluster.annotation, levels = c("pre-meiotic", "early pachytene", "late pachytene"))
markers <- c("Cenpa", "Kit", "H2afx", "Mybl1", "Piwil1", "Ldhc")
p <- VlnPlot(seu, features = markers, pt.size = 0, ncol = 2, assay = "RNA", combine = T, log = F)

pdf("../figures/FigureS8A.pdf", width = 6, height = 10)
p
dev.off()
```

### S8B: transcriptome similarity

```{r}
dge.counts <- readRDS("~/data/data3/jarning/seq-data/z541.ss2/counts.pseudo.bulk.rds")
cellres <- readRDS("~/data/data3/jarning/public/TestisAtlas/data/cache/CellRes2018_30061742.rds")
in.house <- list(
  counts = do.call(cbind, dge.counts)
)
colnames(in.house$counts) <- paste0(rep(c("eP_WT","eP_KO","lP_WT","lP_KO"), each=3), 
                                    rep(c(".r1",".r2",".r3"), 4))
head(in.house$counts)
in.house$tpm <- apply(in.house$counts, 2, function(x) x/sum(x)*1e6)

select.genes <- intersect(rownames(cellres$mean_tpm), rownames(in.house$tpm))
mat <- cbind(cellres$mean_tpm[select.genes, ], in.house$tpm[select.genes, ])
```

```{r, fig.width=8, fig.height=6}
corr <- cor(log2(mat+1), method = "pearson")
pdf("../figures/FigureS8B.pdf", width = 8, height = 6)
pheatmap::pheatmap(corr[9:nrow(corr),9:ncol(corr)], cluster_cols = F, cluster_rows = F, display_numbers = T, number_color = "black")
dev.off()
```

## Figure S9

> 10x single-cell RNA-seq on sorted tetraploid cells from WT and Zfp541-/- testis.

```{r}
library(Seurat)
library(myRutils)
library(tidyverse)

theme_dimplot <- function (base_size = 11, base_family = "", base_line_size = base_size/22, base_rect_size = base_size/22)
{
  half_line <- base_size/2
  theme_bw(base_size = base_size, base_family = base_family,
           base_line_size = base_line_size, base_rect_size = base_rect_size) %+replace%
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.border = element_rect(fill = NA, colour = "black", size = base_line_size),
          panel.grid = element_blank(),
          strip.background = element_blank())
}
```

### S9A: UMAP (10x)
```{r}
seu <- readRDS("~/data/data3/jarning/seq-data/SHIP.10X/seu.rm_doublets.addMS.rds")
cluster.annotation <- read_tsv("../../zhangxinxia_scrna/analysis_SHIPC/clusters.annotations.txt")
seu$annotation <- plyr::mapvalues(seu$seurat_clusters.reorder, cluster.annotation$cluster_index, cluster.annotation$annotation)

seu$orig.ident <- factor(seu$orig.ident, levels = c("WT","Z541"))
seu$genotype <- plyr::mapvalues(seu$orig.ident, c("WT","Z541"), c("WT", "Zfp541(-/-)"))
```

```{r, fig.width=10, fig.height=5}
dims = paste0("UMAP_", 1:2)
meta = c("genotype", "annotation")
data <- FetchData(seu, vars = c(dims, meta))

p <- data %>% 
  ggplot(aes(UMAP_1, UMAP_2)) + 
  geom_point(inherit.aes = F, data = data %>% dplyr::select(-genotype), aes(UMAP_1, UMAP_2), size = 0.1, color="grey", alpha = 0.1) + 
  geom_point(aes(color=annotation), size = .5) + 
  ggsci::scale_color_d3("category20") + 
  guides(color = guide_legend(title = "", override.aes = list(size = 3))) + 
  labs(x="", y="") + 
  facet_wrap(~genotype, nrow = 1) + 
  theme_dimplot(base_size = 20) + 
  theme(strip.text = element_blank())

png("../figures/FigureS9A.png", width = 10, height = 5, unit="in", res=600)
p
dev.off()
```

### S9B: cell counts

```{r, fig.width=8, fig.height=5}
data.use <- data %>% 
  group_by(genotype, annotation) %>% 
  summarise(nCells = n()) %>% 
  ungroup()

data.use <- data.use %>%
  mutate(annotation = factor(annotation, levels = rev(levels(seu$annotation))) )

p <- data.use %>% 
  ggplot(aes(annotation, nCells)) + 
  geom_col(aes(fill=annotation), show.legend = F) + 
  scale_fill_manual(values = rev(ggsci::pal_d3("category20")(15)) ) + 
  coord_flip() + 
  facet_grid(~genotype) + 
  NoLegend() + 
  scale_x_discrete(position = "top") + 
  ylab("Number of cells") + xlab("") + 
  theme_bw(base_size = 20) + 
  theme(axis.line = element_line(color="black"),
        axis.text = element_text(color = "black"),
        plot.title = element_text(hjust = .5, face = "bold"),
        panel.grid = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        legend.background = element_rect(fill=alpha('white', 0))
  )

png("../figures/FigureS9B.png", width = 8, height = 5, units = "in", res = 600)
p
dev.off()
```

### S9C: markers

```{r, fig.width=8, fig.height=5}
p <- DotPlot(seu, group.by = "annotation",
             features = c("Dazl","Nanos3","Ccnd2","Kit",
                          "Stra8","Gm4969","Dmc1","Meiob","Rad51ap2",
                          "Rpl10","Rpl10l","Mybl1","Piwil1","Pou5f2","Ccna1","Tex36","Acrv1",
                          "Prm1","Cstl1")) + 
  labs(x = "", y = "") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"))

pdf("../figures/FigureS9C.pdf", width = 8, height = 5)
p
dev.off()
```

### S9D: integrated UMAP

```{r, fig.width=12, fig.height=5}
data.use <- read_tsv("../../zhangxinxia_scrna/analysis_SHIPC/cell_meta.integrated.tsv")

get_label_pos <- function(data, emb = "tSNE", group.by="ClusterID", label.by="ClusterID") {
  new.data <- data[, c(paste(emb, 1:2, sep = "_"), group.by, label.by)]
  colnames(new.data) <- c("x","y","cluster","label")
  clusters <- names(table(new.data$cluster))
  new.pos <- lapply(clusters, function(i) {
    tmp.data = subset(new.data, cluster == i)
    data.frame(
      x = median(tmp.data$x),
      y = median(tmp.data$y),
      group = tmp.data$cluster[1],
      label = tmp.data$label[1])
  })
  do.call(rbind, new.pos)
}

color.use1 <- ggsci::pal_d3("category20")(15) 
names1 <- levels(seu$annotation)

color.use2 <- readLines("~/data/data3/jarning/project2020/zhangxinxia_scrna/public/20stages.color.txt")
names2 <- names(table(cellres$cell_type))
color.use <- c(color.use1, color.use2)

data.use$cluster <- factor(data.use$cluster,
                           levels = c(paste("10X", names1, sep = "_"),
                                      paste("smart-seq", names2, sep = "_")))

p1 <- ggplot() + 
  geom_point(data = data.use %>% filter(tech == "10X"),
             aes(UMAP_1, UMAP_2, fill=factor(celltype, levels=names1)), size = 3, shape=21, color="black") +
  scale_fill_manual(values = color.use1,
                    breaks = names1) + 
  theme_classic(base_size = 20) + 
  theme(legend.title = element_blank())

lgd <- cowplot::get_legend(p1)

p2 <- ggplot() + 
  geom_point(data = data.use %>% filter(tech == "10X"),
             aes(UMAP_1, UMAP_2, color=cluster), size = 1) + 
  geom_point(data = data.use %>% filter(tech == "smart-seq"), 
             aes(UMAP_1, UMAP_2), size = 3.5, shape=21, fill="black") + 
  geom_point(data = data.use %>% filter(tech == "smart-seq"), 
             aes(UMAP_1, UMAP_2, color=cluster), size = 2.8) + 
  ggrepel::geom_label_repel(data = get_label_pos(data.use %>% filter(tech == "smart-seq"), 
                                                 emb = "UMAP", group.by = "cluster", label.by = "celltype"), 
                            aes(x, y, label = label, color=group), size=4) + 
  scale_color_manual(values = color.use,
                     breaks = levels(data.use$cluster)) +
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

p3 <- ggplot() + 
  geom_point(data = data.use %>% filter(tech == "10X"),
             aes(UMAP_1, UMAP_2, color=cluster), size = 1) + 
  scale_color_manual(values = color.use,
                     breaks = levels(data.use$cluster)) +
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

png("../figures/FigureS9D.png", width = 12, height = 5, units = "in", res = 600)
cowplot::plot_grid(p2,p3,lgd, rel_widths = c(0.4,0.4,0.2), nrow = 1)
dev.off()
```

## Figure S10

> Verification of the DEGs in Zfp541-/- late pachytene spermatocytes identified by Smart-seq2 using 10x scRNA-seq data.

### S10A: compare 10x & ss2 (venn plot)

### S10B: compare 10x & ss2 (volcano plot)

### S10C: DEGs (10x)

```{r}
library(myRutils)
```

```{r}
seu.1 <- subset(seu, annotation %notin% c("D","RS.1","RS.2","RS.3","ES","MI","Diff.Spg1","Diff.Spg2","Diff.Spg3"))
Idents(seu.1) <- seu.1$annotation
```

```{r, fig.width=12, fig.height=5}
genes.choose <- c(
  "Prdm9", ## H3K4me3 writer
  "Gm960","Mei1", ## DSBs
  "Gm4969","Rxra","Ncoa2","Ythdc2", # Meiosis initiate regulator
  "Zcwpw1", # H3K4me3 reader
  "Mcmdc2","Rad51","Hfm1","Rnf212","Msh5","Tex15", # DSB repair and meiotic recombination
  "Syce3","4930447C04Rik", # SC
  "Kit","Uchl1","Dnmt3b","Chd7" # spermatogonia genes
)

p <- DotPlot(seu.1, features = genes.choose, 
             cols = c("blue", "red"), 
             split.by = "orig.ident") + 
  labs(x="", y="") + RotatedAxis() + 
  theme(axis.text.y = element_text(color = c("blue","red")),
        axis.text.x = element_text(face = "italic", angle = 45))

pdf("../figures/Figures10C.pdf", width = 10, height = 5)
p
dev.off()
```

### S10D: GO of down in KO

## Figure S11

> Gene module scores of gene signatures during spermatogenesis along the germ cell developmental trajectory in WT and Zfp541-/- testis.

```{r}
signatures <- c(
  "31P: SSC",
  "31N: Diff.Spg",
  "33N: broad Spg",
  "2P: B.Spg",
  "5N: (pre)Leptotene",
  "44N: Zygotene",
  "38P: MSCI",
  "13N: early Pachytene",
  "39N: late Pachytene",
  "20N: Meiotic Division"
)

names(signatures) <- c("V31P", "V31N", "V33N", "V2P", "V5N", "V44N", "V38P", "V13N", "V39N", "V20N")

length(signatures)
```

```{r, fig.width=16, fig.height=8}
gene_modules = names(signatures)
dims = c("pc.pseudotime")
meta = c("annotation", gene_modules, "genotype")
data <- FetchData(seu, vars = c(dims, meta))
data <- data %>% filter(annotation %notin% c("ES",paste0("RS.", 1:3)))

plot.list <- lapply(gene_modules, function(xx) {
  pts <- c(min(data$pc.pseudotime), median(data$pc.pseudotime), max(data$pc.pseudotime))
  data.use <- flattern_imputated_data(data, bin.width = 10, value = xx)
  ggplot() +
    geom_line(data = data.use, aes(pc.pseudotime, median.exp, color=genotype, group=genotype), size = 0.5) +
    geom_jitter(data = data, aes(pc.pseudotime, -0.3, fill = annotation), show.legend = F,
                stroke = 0, color="white", shape=21, height = 0.02, size = 1, alpha = .5) +
    ggsci::scale_fill_d3("category20") + 
    geom_vline(xintercept = -22540, linetype="dashed", color="blue") +
    scale_x_continuous(breaks = pts, labels = seq(0,100,50)) + 
    guides(color = guide_legend(title = "", override.aes = list(size = 1.5))) + 
    labs(x="Pseudotime", y="Module Score", title = signatures[xx]) + 
    theme(legend.position = c(0.8, 0.95))
})

png("../figures/FigureS11.png", width = 16, height = 8, unit="in", res=600)
plot.list <- lapply(plot.list, function(p) p + theme(legend.position = "none"))
cowplot::plot_grid(plotlist = plot.list, nrow = 2)
dev.off()
```
